---
title: "Extra atlases for ggseg-package"
author: "Athanasia Mowinckel & Didac Vidal Pineiro"
output: github_document
---

  [![Travis build status](https://travis-ci.com/LCBC-UiO/ggsegExtra?branch=master)](https://travis-ci.com/LCBC-UiO/ggsegExtra)
  [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/LCBC-UiO/ggsegExtra?branch=master&svg=true)](https://ci.appveyor.com/project/LCBC-UiO/ggsegExtra)
  [![Coverage status](https://codecov.io/gh/LCBC-UiO/ggsegExtra/branch/master/graph/badge.svg)](https://codecov.io/github/LCBC-UiO/ggsegExtra?branch=master)
  
```{r setup, include=FALSE}
library(ggseg)
knitr::opts_chunk$set(echo = TRUE)
```

This package contains extra atlases for the use in the [ggseg]() plotting package in R.
## Atlases
There are currently three atlases available in the package:  

1. `midsagittal` - mid-sagittal slice showing the cerebellum, 3^rd^ and 4^th^ ventricles, corpus callosum etc.
2. `glasser` - HCP cortical segmentation (Glasser et al. 2015)  
3. `jhu`- JHU white-matter tractography atlas  

We are working on creating a detailed description in the wiki on how to create and contribute atlases to the package.
The `ggseg` function already allows you to provide it with a data.frame of a custom atlas if you have it, but is must
correspond to certain specifications to work.

Please see the [wiki](https://github.com/LCBC-UiO/ggseg/wiki/Creating-and-contributing-atlases) for information on adding atlases, or inspect the included datasets for requirements. 
If anything is unclear in the wiki, give us a shout out in the issues!


## Use
You will need to use the `ggseg` package to use these atlases, or wrangle them your self.
You can access it [here](https://lcbc-uio.github.io/ggseg/ggseg.html), or via R:

```{r eval=F}
devtools::install_github("LCBC-UiO/ggseg")
```


### Report bugs or requests  
Don't hesitate to ask for support using [github issues](https://github.com/LCBC-UiO/ggseg/issues), or requesting new atlases. 
While we would love getting help in creating new atlases, you may also request atlases through the issues, and we will try to get to it. 

### Included atlases and palettes
```{r glasser, fig.height=12, echo=F}
ggseg(atlas=glasser,position="stacked", mapping=aes(fill=area)) +
  scale_fill_brain("glasser") +
  ggtitle("HCP cortical segmentation (glasser)") +
  theme(legend.position = "bottom")+
  guides(fill=guide_legend(ncol=8))
```

```{r midsagittal, echo=F}
ggseg(atlas=midsagittal, mapping=aes(fill=area)) +
  scale_fill_brain("midsagittal") +
  ggtitle("Mid-sagittal subdortical, cerebellum, & corpus callosum")
```

```{r jhu, echo=F}
ggseg(atlas=jhu, mapping=aes(fill=area)) +
  scale_fill_brain("jhu") +
  ggtitle("JHU white-matter tractography atlas") +
  theme(legend.position = "bottom")+
  guides(fill=guide_legend(ncol=2))
```



# Funding
This tool is partly funded by:  

**EU Horizon 2020 Grant:** Healthy minds 0-100 years: Optimising the use of European brain imaging cohorts (Lifebrain).

**Grant agreement number:** 732592.

**Call:** Societal challenges: Health, demographic change and well-being
