---
title: "System Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{System Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: load-pkg
library(ggsegExtra)
```

## FreeSurfer

Creating custom atlases requires [FreeSurfer](https://surfer.nmr.mgh.harvard.edu/fswiki/DownloadAndInstall) to be installed and configured.

On macOS, you also need:

- [XQuartz](https://www.xquartz.org/)
- [Xcode](https://developer.apple.com/xcode/) command line tools

## Plotly screenshots

For taking screenshots of 3D plotly brain visualizations, ggsegExtra uses the webshot2 package, which requires Google Chrome or Chromium.

```r
install.packages("webshot2")
```

Chrome/Chromium is typically already installed. If not, webshot2 will attempt to download a suitable version automatically.

## Parallel processing

Atlas creation can be slow, especially for cortical atlases with many regions.
ggsegExtra uses the [furrr](https://furrr.futureverse.org/) package for parallel processing.
By default, processing runs sequentially.
To enable parallel processing, set up a `future` plan before running atlas creation functions:
```r
library(future)

# Use all available cores
plan(multisession)

# Or specify number of workers
plan(multisession, workers = 4)
```

To return to sequential processing:

```r
plan(sequential)
```

## Progress bars

ggsegExtra uses the [progressr](https://progressr.futureverse.org/) package to report progress during long-running operations.
Progress reporting is disabled by default.
To enable progress bars, wrap your code in `with_progress()` or set a global handler:

```r
library(progressr)

# Option 1: Wrap specific code
with_progress({
  atlas <- make_ggseg3d_2_ggseg(...)
})

# Option 2: Enable globally for the session
handlers(global = TRUE)
```

You can customize the progress bar style:

```r
# Use a simple text progress bar
handlers("txtprogressbar")

# Use cli-style progress (recommended)
handlers("cli")
```

